Mortality Trends Among Shelter Residents in Toronto: A Longitudinal Study from 2007 to 2024

### 2024-09-27 starter_folder-main/paper/paper.qmd
The introduction done.

---
title: "Mortality Trends Among Shelter Residents in Toronto: A Longitudinal Study from 2007 to 2024"
author: 
  - Fangning Zhang
thanks: "Code and data are available at: https://github.com/FangningZhang81/Deaths-of-Shelter-Residents."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
#hide the code in the PDF

@@

@@

The remainder of this paper is structured as follows. @sec-data....

@@

## Overview
The dataset used in this study is titled "Deaths of Shelter Residents". The dataset is published by the Toronto Shelter and Support Services Department and is obtained from OpenDataToronto (@opendatatorotno), which is accurate and reliable. It covers the deaths of shelter residents from 2007 to the present. Variables such as the total number of decedents and the breakdown of deaths by gender (male, female, transgender/non-binary/two-spirit) are included. All death counts are reported on a monthly basis and is still updated monthly. For privacy and ethical reasons, the dataset does not contain any personally identifiable information.

@@

# Discussion

@@

One limitation of this study is that the original dataset does not account for Transgender/Non-binary/Two-Spirit individuals before 2020, which reduces the accuracy of gender-specific analyses and raises ethical concerns. Additionally, the number of deaths may depend on the number of shelter residents, but this information is not included in the dataset, making it impossible to calculate mortality rates. Furthermore, changes in shelter admission conditions and other policies, which could impact the number of deaths, are not discussed in this paper.

## Weaknesses and next steps

Weaknesses and next steps should also be included.

>>>>

---
title: "Mortality Trends Among Shelter Residents in Toronto: A Longitudinal Study from 2007 to 2024"
author: 
  - Fangning Zhang
thanks: "Code and data are available at: https://github.com/FangningZhang81/Deaths-of-Shelter-Residents."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
#hide the code in the PDF

@@

Shelter residents are one of the most vulnerable groups in urban areas, facing numerous challenges such as housing instability, health issues, and social exclusion. Numerous studies have shown that shelter residents are at a significantly higher risk of premature death compared to the general population due to factors such as poor health, lack of access to medical care, and exposure to unsafe environments(@hwang2009mortality). Understanding the mortality of shelter residents is crucial for formulating effective social policies, improving shelter services and also improving the health and well-being of shelter residents.

Although many existing researches highlight the higher mortality rates among the homeless population, there is still a lack of analysis specifically targeting shelter residents. In this study, a comprehensive analysis of mortality data is made among shelter residents in Toronto, focusing on the following key aspects. First, we analyze the overall trend in the number of deaths over the 16-year period from 2007 to 2023. Second, we explored gender differences in mortality. Third, we analyzed the monthly distribution of deaths to understand whether different climates or seasons have varying impacts on shelter residents.

The analysis revealed a concerning upward trend in the number of deaths starting from 2019, despite a slight decline in the following two years. Male deaths number have consistently been higher than female deaths, which is more and more significant after 2019. Additionally, there was an increase in deaths during the winter months, suggesting a link between extreme weather conditions and mortality risk may exist. This may consistent with previous research indicating that bad weather conditions and overcrowding in shelters can exacerbate health risks for homeless individuals (@vuillermoz2016mortality).

The rise in the number of deaths starting in 2019 worths attention. However, this increase does not necessarily indicate worsening shelter conditions or declining health among residents. The rise could be due to an increase in the number of shelter residents, changes in admission policies, or external factors such as the impact of COVID-19 and other social and public health issues (@ramin2009health). This is discussed in detail in the discussion section.

@@

The remainder of this paper is structured as follows:@sec-data describes the data sources, tools, cleaning process, visualization, result and related datasets; and @sec-discussion summarizes the key findings, interprets the results, discusses limitations, and suggests directions for future research.

@@

## Data Source and Measurement
The dataset used in this study is titled "Deaths of Shelter Residents". This dataset is published by the Toronto Shelter and Support Services Department and is obtained from OpenDataToronto (@opendatatorotno). The Toronto Shelter and Support Services Division has collected this information since 2007 and conducts an annual review of the data. Besides, the death of shelter residents data is a subset of data shared by Toronto Public Health (TPH) on its Deaths of People Experiencing Homelessness dashboard. The TPH dashboard includes figures on deaths of individuals experiencing homelessness staying in shelters, as well as people living outdoors or at other locations. Therefore, the dataset is accurate and reliable. The dataset covers the deaths of shelter residents from 2007 to the present. Variables such as the total number of decedents and the breakdown of deaths by gender (male, female, transgender/non-binary/two-spirit) are included. All death counts are reported on a monthly basis and is still updated monthly. For privacy and ethical reasons, the dataset does not contain any personally identifiable information.

@@

# Discussion {#sec-discussion}

@@

One limitation of this study is that the original dataset does not account for Transgender/Non-binary/Two-Spirit individuals before 2020, which reduces the accuracy of gender-specific analyses and raises ethical concerns. Additionally, the number of deaths may depend on the number of shelter residents, but this information is not included in the dataset, making it impossible to calculate mortality rates. The rapid growth of the death number after 2019 may be influenced by the increasing number of shelters or Covid-19, etc. Furthermore, changes in shelter admission conditions and other policies, which could impact the number of deaths, are also not discussed in this paper.

## Suggestions for Future Research

Future research should explore external factors such as economic conditions and the availability of healthcare resources that may further influence the number of deaths among shelter residents. It is recommended to utilize more diverse datasets, such as daily shelter occupancy rates which is mentioned earlier and other health-related data, to gain a deeper understanding of the health and mortality trends of shelter residents. These approaches would offer a deeper understanding of the complex factors influencing the vulnerable population mortality.
